Constraining the Interiors of Asteroids Through Close Encounters

### 2022-07-29 paper/appendices.tex
Ready to push new runs to SC & responded to some revisions

\subsection{Observational uncertainty}
\label{sec:scan-uncertainty}
Two parameters, $\sigma_\theta$ and $\sigma_P$, govern the observational uncertainty of the data set. These parameters are defined in section \ref{sec:uncertainty}; $\sigma_\theta$ represents the standard deviation of the angle between the true spin pole and the observed spin pole, while $\sigma_P$ represents the standard deviation of the rotational period. Rather than explore the full space spanned by these two values, we fix one and allow the other to vary to better assess whether uncertainty in spin pole or uncertainty in period more strongly affects uncertainty. This dependence is displayed in figure \ref{fig:scan-theta}. More precisely, we can ask if some error $\bm {\delta}_\omega$ is added to angular velocity $\omega$, does it affect results more strongly if it is parallel to $\omega$ (affects the period) or perpendicular (affects the spin pole direction)?

@@

Comparing the $\sigma_P / P$ (fractional change in period) and $\sigma_\theta$ (spin pole angle) columns of figure \ref{fig:scan-observational}, one can see that a given value of $\sigma_\theta$ contributes a much smaller moment uncertainty than the same value of $\sigma_P/ P$. This is also visible in figure \ref{fig:net-uncertainty} for the lumpy model. In other words, if $\bm {\delta}_\omega \perp \bm \omega$ using the symbols defined above, then $\delta$ can be large. But if $\delta = |\bm{\delta}_\omega| / |\bm \omega|$, then $\delta$ must be very small. Period precision is therefore more vital than spin pole direction precision when it comes to decreasing density distribution uncertainties.

>>>>

\subsection{Observational uncertainty}
\label{sec:scan-uncertainty}
Two parameters, $\sigma_\theta$ and $\sigma_P$, govern the observational uncertainty of the data set. These parameters are defined in section \ref{sec:uncertainty}; $\sigma_\theta$ represents the standard deviation of the angle between the true spin pole and the observed spin pole, while $\sigma_P$ represents the standard deviation of the rotational period. Rather than explore the full space spanned by these two values, we fix one and allow the other to vary to better assess whether uncertainty in spin pole or uncertainty in period more strongly affects uncertainty. This dependence is displayed in figure \ref{fig:scan-theta}. More precisely, we can ask if some error $\bm {\delta}_\omega$ is added to angular velocity $\bm \omega$, does it affect results more strongly if it is parallel to $\bm \omega$ (affects the period) or perpendicular (affects the spin pole direction)?

@@

Comparing the $\sigma_P / P$ (fractional change in period) and $\sigma_\theta$ (spin pole angle) columns of figure \ref{fig:scan-observational}, one can see that a given value of $\sigma_\theta$ contributes a much smaller moment uncertainty than the same value of $\sigma_P/ P$. This is also visible in figure \ref{fig:net-uncertainty} for the lumpy model. In other words, if $\bm {\delta}_\omega \perp \bm \omega$ using the symbols defined above, then $\delta$ can be large. But if $\bm{\delta}_\omega  \parallel \bm \omega$, then $\delta$ must be very small. Period precision is therefore more vital than spin pole direction precision when it comes to decreasing density distribution uncertainties.

### 2022-07-29 paper/main.tex
Ready to push new runs to SC & responded to some revisions

%%%%% AUTHORS - PLACE YOUR OWN COMMANDS HERE %%%%%
\newcommand{\jtd}[1]{ {\bf{\color{red} JTD: #1}} }
\newcommand{\unit}[1]{\bm{\hat{#1}}}

@@

Angular velocity perturbations have been observed and used to extract asteroid properties in several cases, including for the 2013 encounter of (367943) Duende with Earth \cite{MOSKOVITZ2020113519, benson2020spin}, and asteroid binaries (3905) Doppler and (617) Patroclus \cite{DESCAMPS2020113726, BERTHIER2020113990}. Orbital and physical properties, including MOI ratios have also been extracted for asteroid 99942 Apophis, discovered on June 19, 2004 by R. A. Tucker, D. J. Tholen, and F. Bernardi \cite{giorgini2005recent, giorgini2008predicting, smalley20052004}, and on target to encounter Earth in 2029 \cite{yu2014numerical, hirabayashi2021finite, valvano2022apophis, Lee2022Apophis}. However, density distribution features beyond the moment of inertia (MOI) ratios have not yet been extracted for any asteroid encounters. More research is needed to study in what cases these effects are observable, and what factors generally inhibit observation of these new features. It seems pivotal to augment previous work on the affect of tidal torque on Apophis' angular velocity in particular \cite{souchay2014rotational, souchay2018changes} so that upcoming observations may constrain these properties and thus improve our predictions.

@@

The only properties of an asteroid's density moments that affect tidal torque interactions are its ``density moments,'' defined here as
\begin{equation}
  K_{\ell m} = \frac{a_\mathcal{A}^{2-\ell}}{I_\mathcal{A}} \int_\mathcal{A} d^3 r \rho_\mathcal{A}(\bm r) R_{\ell m}(\bm r).
  \label{eqn:klm}
\end{equation}
These are complex, unitless quantities. $\rho_\mathcal{A}(\bm r)$ is the asteroid density distribution and $R_{\ell m}$ are the regular solid spherical harmonics (see appendix \ref{app:eom} for details). The integral is computed over the entire asteroid mass, denoted $\mathcal{A}$. $I_\mathcal{A}$ denotes a MOI scale defined as 
\begin{equation}
  I_\mathcal{A} = \int_\mathcal{A} d^3 r \rho_\mathcal{A}(\bm r) r^2
  \label{eqn:ia}
\end{equation}
while $a_\mathcal{A}$ is the length scale
\begin{equation}
  a_\mathcal{A}^2 = \frac{1}{V_\mathcal{A}} \int_\mathcal{A} d^3 r r^2
  \label{eqn:aa}
\end{equation}
where $V_\mathcal{A}$ is the asteroid volume. We call these MOI and length scales in part because they obey $I_\mathcal{A} = \mu_\mathcal{A} a_\mathcal{A}^2$ for uniform asteroids where $\mu_\mathcal{A}$ is the mass of the asteroid. Note that $a_\mathcal{A}$ is a function only of the surface of the asteroid, so that $a_\mathcal{A}$ is known if the surface is observed.

@@

Since it is the angular acceleration of the asteroid that is observable, rather than the torque applied, we also compute the MOI of the asteroid around the principal axes: 
\begin{equation}
  \begin{split}
    I_x &= \frac{2}{3} I_\mathcal{A} \parens{K_{20} - 6 K_{22} + 1}\\
    I_y &= \frac{2}{3} I_\mathcal{A} \parens{K_{20} + 6 K_{22} + 1}\\
    I_z &= \frac{2}{3} I_\mathcal{A} \parens{-2K_{20} + 1}.
  \end{split}
  \label{eqn:moi}
\end{equation}

The angular acceleration of the asteroid is proportional to $\bm \tau / I$ such that the $I_\mathcal{A}$ factor of equation \ref{eqn:tidal-torque} is canceled out. Thus, the observables do not depend explicitly on $I_\mathcal{A}$.

@@

We assume for simplicity that the asteroid is initially not tumbling, though this assumption can be relaxed. We assume a short-axis rotation mode such that energy of the asteroid's rotation is minimized. This sets two of the Euler angles equal to zer0 initially, so that only $\gamma_0$ is necessary to provide initial data for the simulation.

@@

\subsection{Extracting density moments from spin data}
\label{sec:fit}
Given synthetic data, an Affine Invariant MCMC Ensemble sampler was used to generate Posterior Probability Distributions (PPDs) from flat priors. We use the Python implementation \texttt{emcee} \cite{foreman2013emcee}. Our parameters were $\gamma_0$, $K_{20}$, $K_{22}$, and $K_{3m}$ (10 in total), bounded by $|\gamma_0| < \pi/4$, and the $K_{2 m}$ bounds given in equation \ref{eqn:parameter-bounds}. We also use $|K_{3m}| < 1$. In general, spin data is most sensitive to $\gamma_0$ and $K_{2m}$, which we call the ``first-order parameters.'' We call $K_{3m}$ the ``second-order parameters.''

The MCMC was determined to converge when the fractional change in autocorrelation time (computed every 100 iterations) was one percent, and the number of iterations already computed was more than 100 times the autocorrelation time. The MCMC fit also was set to terminate if more than $10^5$ iterations were run, but this only occurred for fits computed for asteroid encounters in which degeneracies were present. This degeneracy could be removed by only fitting $K_{2m}$ instead. About $10^4$ iterations was often sufficient.

@@

We rotate the known surface of the asteroid to a new frame called the ``hybrid frame.'' The hybrid frame is co-located with the body-fixed frame, but its orientation is known with respect to the inertial frame. (It has $\bm \Omega_0 \parallel \unit z_\text{hybrid}$ with third Euler angle $\gamma = 0$.) The hybrid frame differs from the body-fixed frame only by a rotation around $\unit z_\text{hybrid}=\unit z_\text{body-fixed}$ of $\gamma_0$. Such rotations affect density moments via 
\begin{equation}
  K_{\ell m}^\mathrm{hybrid} = e^{-im\gamma_0}K_{\ell m}^\mathrm{body-fixed}.
  \label{eqn:body-fixed-to-hybrid}
\end{equation}
by equation \ref{eqn:ylm-rotation}. Thus, values and  uncertainties on $K_{\ell m}^\mathrm{body-fixed}$ and $\gamma_0$ (obtained from the encounter data) can be translated into values and uncertainties on $K_{\ell m}^\mathrm{hybrid}$. Henceforth, we will operate only in the hybrid frame and suppress the label.

@@

  \caption{Cross-sectional slices of the density distributions extracted via the finite element model for the asymmetric (\textit{top two rows}) and symmetric (\textit{bottom two rows}) reference asteroids. The finite element model (\textit{first and third rows}) and the lumpy model (\textit{second and fourth rows}) are employed. From left to right, the densities (divided by the average density), deviations from the true density, uncertainties, and significance of the deviations are plotted. These figures are available in animated form in appendix figure \ref{fig:animated-uniform}. Extracted densities are generally within 10\% of the truth.}
  \label{fig:den-uniform}
\end{figure*}

@@

The lumpy model also yields distributions which are consistent with the extracted moments and consistent with uniform. These distributions have much lower uncertainty than the finite element distributions (maximum uncertainty on the order of 1\% of the local density or less) due to its few degrees of freedom. With only one lump, the location of the known center of mass requires that the lump lie at the centroid of the asteroid center with mass close to zero and unconstrained radius. Hence, the uncertainty of regions far from the asteroid is typically very small, while regions close to the center are more likely to be contained inside a lump.

@@

\begin{figure*}
  \centering
  \includegraphics[width=\linewidth]{figs/unc-scan.pdf}
  \caption{Median ratio of density uncertainty to density computed over the reference asteroid density distribution as a function of physical / observational encounter properties (\textit{top / bottom}). Density distributions from both the lumpy and finite element models are used. The uncertainty thresholds (\textit{red lines}) are shown and the properties that exceed these thresholds are shaded. Te vertical black dotted lines are the property values of the reference asteroid. The figures illustrate the marked difference in density distributions that results from using different models.}
  \label{fig:net-uncertainty}
\end{figure*}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.7\columnwidth]{figs/oblateness.pdf}
  \vfill
  \includegraphics[width=0.7\columnwidth]{figs/oblateness-differ.pdf}
  \caption{\textit{Top}: 1 and 2$\sigma$ confidence intervals for the first-order parameter PPDs as a function of oblateness $\epsilon$ . Linear best-fitting lines to $\sigma(K_{2m})$ (black, dotted) are plotted. \textit{Bottom}: The difference between PPD means extracted from a zero-oblateness model and the true parameters given data with true oblateness $\epsilon_\text{true} \neq 0$. Also shown in both figures are the oblatenesses of reference Solar System bodies. Moment uncertainty depends little on oblateness, but the best-fitting parameter estimates are affected enough by oblateness that oblateness must still be modelled.}
  \label{fig:scan-oblateness}
\end{figure}

Figure \ref{fig:scan-oblateness} displays moment uncertainty $\sigma(K_{2 m})$ of the first-order parameters as a function of $\epsilon$ across a reasonable range of central body oblatenesses based on those of Solar System planets \cite{paterLissauer2015}. It also shows linear best-fitting curves for moment uncertainty as a function of oblateness. Almost no dependence of $\sigma(K_{\ell m})$ on oblateness $\epsilon$ is apparent, although moment uncertainty does measurably decrease for oblate central bodies.

>>>>

%%%%% AUTHORS - PLACE YOUR OWN COMMANDS HERE %%%%%
\newcommand{\jtd}[1]{ {\bf{\color{red} JTD: #1}} }
\renewcommand{\unit}[1]{\bm{\hat{#1}}}

@@

Angular velocity perturbations have been observed and used to extract asteroid properties in several cases, including for the 2013 encounter of (367943) Duende with Earth \cite{MOSKOVITZ2020113519, benson2020spin}, and asteroid binaries (3905) Doppler and (617) Patroclus \cite{DESCAMPS2020113726, BERTHIER2020113990}. Orbital and physical properties, including moment of inertia (MOI) ratios have also been extracted for asteroid 99942 Apophis, discovered on June 19, 2004 by R. A. Tucker, D. J. Tholen, and F. Bernardi \cite{giorgini2005recent, giorgini2008predicting, smalley20052004}, and on target to encounter Earth in 2029 \cite{yu2014numerical, hirabayashi2021finite, valvano2022apophis, Lee2022Apophis}. However, density distribution features beyond the MOI ratios have not yet been extracted for any asteroid encounters. More research is needed to study in what cases these effects are observable, and what factors generally inhibit observation of these new features. It seems pivotal to augment previous work on the affect of tidal torque on Apophis' angular velocity in particular \cite{souchay2014rotational, souchay2018changes} so that upcoming observations may constrain these properties and thus improve our predictions.

@@

The only properties of an asteroid's density moments that affect tidal torque interactions are its ``density moments,'' defined here as
\begin{equation}
  K_{\ell m} = \frac{a_\mathcal{A}^{2-\ell}}{I_\mathcal{A}} \int_\mathcal{A} d^3 r \rho_\mathcal{A}(\bm r) R_{\ell m}(\bm r).
  \label{eqn:klm}
\end{equation}
These are complex, unitless quantities. $\rho_\mathcal{A}(\bm r)$ is the asteroid density distribution and $R_{\ell m}$ are the regular solid spherical harmonics (see appendix \ref{app:eom} for details). The integral is computed over the entire asteroid mass, denoted $\mathcal{A}$, and $d^3 r$ indicates the three-dimensional volume element throughout the paper. $I_\mathcal{A}$ denotes a MOI scale defined as 
\begin{equation}
  I_\mathcal{A} = \int_\mathcal{A} d^3 r \rho_\mathcal{A}(\bm r) r^2
  \label{eqn:ia}
\end{equation}
while $a_\mathcal{A}$ is the length scale
\begin{equation}
  a_\mathcal{A}^2 = \frac{1}{V_\mathcal{A}} \int_\mathcal{A} d^3 r r^2
  \label{eqn:aa}
\end{equation}
where $V_\mathcal{A}$ is the asteroid volume. We call these MOI and length scales in part because they obey $I_\mathcal{A} = \mu_\mathcal{A} a_\mathcal{A}^2$ for uniform asteroids where $\mu_\mathcal{A}$ is the mass of the asteroid. Note that $a_\mathcal{A}$ is a function only of the surface of the asteroid, so that $a_\mathcal{A}$ is known if the surface is observed.

@@

Since it is the angular acceleration of the asteroid that is observable, rather than the torque applied, we also compute the MOI of the asteroid around the principal axes: 
\begin{equation}
  \begin{split}
    I_x &= \frac{2}{3} I_\mathcal{A} \parens{K_{20} - 6 K_{22} + 1}\\
    I_y &= \frac{2}{3} I_\mathcal{A} \parens{K_{20} + 6 K_{22} + 1}\\
    I_z &= \frac{2}{3} I_\mathcal{A} \parens{-2K_{20} + 1}.
  \end{split}
  \label{eqn:moi}
\end{equation}
Note that all moments of inertia are proportional to $I_\mathcal{A}$. Equation \ref{eqn:tidal-torque} indicated that tidal torque was also proportional to $I_\mathcal{A}$. The Euler equations, giving angular acceleration in terms of moment of inertia and torque (equation \ref{eqn:omega-eom}), show that the angular acceleration of the asteroid is independent $I_\mathcal{A}$. Thus, the observables do not depend explicitly on $I_\mathcal{A}$.

@@

We assume for simplicity that the asteroid is initially not tumbling, though this assumption can be relaxed. We assume a short-axis rotation mode such that energy of the asteroid's rotation is minimized. This sets two of the Euler angles equal to zero initially, so that only $\gamma_0$ is necessary to provide initial data for the simulation.

@@

\subsection{Extracting density moments from spin data}
\label{sec:fit}
Given synthetic data, an Affine Invariant MCMC Ensemble sampler was used to generate Posterior Probability Distributions (PPDs) from flat priors. We use the Python implementation \texttt{emcee} \cite{foreman2013emcee}. Our parameters were $\gamma_0$, $K_{20}$, $K_{22}$, and $K_{3m}$ (10 in total), bounded by $|\gamma_0| < \pi/4$, and the $K_{2 m}$ bounds given in equation \ref{eqn:parameter-bounds}. Computing $K_{3m}$ for many positive density distributions revealed a $|K_{3m}| \lesssim 0.01$, so we use more conservative bounds on $K_{3 m}$ of $|K_{3 m}| < 1$. In general, spin data is most sensitive to $\gamma_0$ and $K_{2m}$, which we call the ``first-order parameters.'' We call $K_{3m}$ the ``second-order parameters.''

The MCMC was determined to converge when the fractional change in autocorrelation time (computed every 100 iterations) was one percent, and the number of iterations already computed was more than 100 times the autocorrelation time. The MCMC fit also was set to terminate if more than $10^5$ iterations were run, but this only occurred for fits computed for asteroid encounters in which degeneracies were present, for example when the asteroid has rotational symmetry. About $10^4$ iterations was often sufficient.

@@

To compare the asteroid surface and extracted density moments in the same coordinate system, we define a new frame called the ``hybrid frame.'' The hybrid frame is co-located with the body-fixed frame, but its orientation is known with respect to the inertial frame. (It has $\bm \Omega_0 \parallel \unit z_\text{hybrid}$ with third Euler angle $\gamma = 0$.) The hybrid frame differs from the body-fixed frame only by a rotation around $\unit z_\text{hybrid}=\unit z_\text{body-fixed}$ of $\gamma_0$. Such rotations affect density moments via 
\begin{equation}
  K_{\ell m}^\mathrm{hybrid} = e^{-im\gamma_0}K_{\ell m}^\mathrm{body-fixed}.
  \label{eqn:body-fixed-to-hybrid}
\end{equation}
by equation \ref{eqn:ylm-rotation}. Thus, values and  uncertainties on $K_{\ell m}^\mathrm{body-fixed}$ and $\gamma_0$ (obtained from the encounter data) can be translated into values and uncertainties on $K_{\ell m}^\mathrm{hybrid}$. Henceforth, we will operate only in the hybrid frame and suppress the label.

@@

  \caption{Cross-sectional slices of the density distributions extracted via the finite element model for the asymmetric (\textit{top two rows}) and symmetric (\textit{bottom two rows}) uniform-density reference asteroids. The finite element model (\textit{first and third rows}) and the lumpy model (\textit{second and fourth rows}) are employed. From left to right, the densities (divided by the average density), deviations from the true density, uncertainties, and significance of the deviations are plotted. These figures are available in animated form in appendix figure \ref{fig:animated-uniform}. Extracted densities are generally within 10\% of the truth.}
  \label{fig:den-uniform}
\end{figure*}

@@

The lumpy model also yields distributions which are consistent with the extracted moments and consistent with uniform. These distributions have much lower uncertainty than the finite element distributions (maximum uncertainty on the order of 1\% of the local density or less) due to its few degrees of freedom. With only one lump and with asteroids where the surface's centroid is the asteroid's center of mass (such as these uniform asteroids), the lump must lie at the centroid with mass close to zero and unconstrained radius. Hence, the uncertainty of regions far from the asteroid is typically very small, while regions close to the center are more likely to be contained inside a lump and hence have more density uncertainty.

@@

\begin{figure*}
  \centering
  \includegraphics[width=\linewidth]{figs/unc-scan.pdf}
  \caption{Median ratio of density uncertainty to density computed over the reference asteroid density distribution as a function of physical / observational encounter properties (\textit{top / bottom}). Density distributions from both the lumpy and finite element models are used. The uncertainty thresholds (\textit{red lines}) are shown and the properties that exceed these thresholds are shaded. The vertical black dotted lines are the property values of the reference asteroid. The figures illustrate the marked difference in density distributions that results from using different models.}
  \label{fig:net-uncertainty}
\end{figure*}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.7\columnwidth]{figs/oblateness.pdf}
  \vfill
  \includegraphics[width=0.7\columnwidth]{figs/oblateness-differ.pdf}
  \caption{\textit{Top}: 1 and 2$\sigma$ confidence intervals for the first-order parameter PPDs as a function of oblateness $\epsilon$. All other parameters, including the central body radius, are kept constant. Linear best-fitting lines to $\sigma(K_{2m})$ (black, dotted) are plotted. \textit{Bottom}: The difference between PPD means extracted from a zero-oblateness model and the true parameters given data with true oblateness $\epsilon_\text{true} \neq 0$. Also shown in both figures are the oblatenesses of reference Solar System bodies. Moment uncertainty depends little on oblateness, but the best-fitting parameter estimates are affected enough by oblateness that oblateness must still be modelled.}
  \label{fig:scan-oblateness}
\end{figure}

Figure \ref{fig:scan-oblateness} displays moment uncertainty $\sigma(K_{2 m})$ of the first-order parameters as a function of $\epsilon$ across a reasonable range of central body oblatenesses based on those of Solar System planets \cite{paterLissauer2015}. It also shows linear best-fitting curves for moment uncertainty as a function of oblateness. All other parameters, include $I_\mathcal{B}$ which parametrizes the central body radius, are kept constant. Almost no dependence of $\sigma(K_{\ell m})$ on oblateness $\epsilon$ is apparent, although moment uncertainty does measurably decrease for oblate central bodies.
